---
title: "data prep"
format: html
---

```{r, echo = FALSE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE
)
options(scipen = 999, digits = 4)
knitr::knit_hooks$set(
  inline = function(x) {
    if (is.numeric(x)) {
      return(prettyNum(x, big.mark = ","))
    } else{
      return(x)
    }
  }
)
```

```{r, echo=FALSE,message = FALSE, warning = FALSE}
library(here)
library(DBI)
library(RPostgres)
library(tidyverse)
library(networkD3)
library(htmlwidgets)
library(htmltools)
library(sankeyD3) # Not on CRAN, but here <https://github.com/fbreitwieser/sankeyD3>
library(scales)
library(ggsci)
library(cowplot)
library(roadoi)
library(rcrossref)
```

# SQL Query to create shared corpus table in KB infrastructure

``` sql 
create table unigsdoerner.scp_shared_corpus as(
select 
scp04.item_id as scp04_item_id,
scp04.doi as scp04_doi,
scp04.oa_status as scp04_oa_status,
scp07.oa_status as scp07_oa_status,
scp10.oa_status as scp10_oa_status
from scp_b_202404.items scp04
left join scp_b_202407.items scp07 on
scp04.item_id = scp07.item_id 
left join scp_b_202410.items scp10 on
scp04.item_id = scp10.item_id 
where scp04.pubyear between 2019 and 2023
);
```

# Load data from shared corpus table

```{sql, connection=kb_con, output.var="scp_shared_corpus"}
select * from unigsdoerner.scp_shared_corpus
```

```{r}
write_csv(scp_shared_corpus, here("scp_shared_corpus.csv"))
```

# Datamanipulation of shared corpus table

- Replace NA values with none
- Remove curly brackets and quotation marks
- Remove rows that only contain all in curly brackets
- Remove "all" and "repository" in all rows
- Remove rows that only contain "all" or "repository"


```{r}
scp_sc_prep <- scp_shared_corpus %>%
  mutate(scp04_oa_status = replace(scp04_oa_status, is.na(scp04_oa_status), "none")) %>%
  mutate(scp07_oa_status = replace(scp07_oa_status, is.na(scp07_oa_status), "none")) %>%
  mutate(scp10_oa_status = replace(scp10_oa_status, is.na(scp10_oa_status), "none")) %>%
  mutate(scp04_oa_status = gsub("\\{\"|\"\\}", "",scp04_oa_status)) %>%
  mutate(scp07_oa_status = gsub("\\{\"|\"\\}", "",scp07_oa_status)) %>%
  mutate(scp10_oa_status = gsub("\\{\"|\"\\}", "",scp10_oa_status)) %>%
  filter(!grepl("^\\{all\\}$",scp04_oa_status)) %>%
  filter(!grepl("^\\{all\\}$",scp07_oa_status)) %>%
  filter(!grepl("^\\{all\\}$",scp10_oa_status)) %>%
  filter(!grepl("^(all repository)$", scp04_oa_status)) %>%
  filter(!grepl("^(all repository)$", scp07_oa_status)) %>%
  filter(!grepl("^(all repository)$", scp10_oa_status)) %>%
  mutate(scp04_oa_status = str_replace(scp04_oa_status, "^all ", "")) %>%
  mutate(scp07_oa_status = str_replace(scp07_oa_status, "^all ", "")) %>%
  mutate(scp10_oa_status = str_replace(scp10_oa_status, "^all ", "")) %>%
  mutate(scp04_oa_status = str_replace(scp04_oa_status, "repository ", "")) %>%
  mutate(scp07_oa_status = str_replace(scp07_oa_status, "repository ", "")) %>%
  mutate(scp10_oa_status = str_replace(scp10_oa_status, "repository ", "")) %>%
  mutate(scp04_oa_status = str_replace(scp04_oa_status, " repository$", "")) %>%
  mutate(scp07_oa_status = str_replace(scp07_oa_status, " repository$", "")) %>%
  mutate(scp10_oa_status = str_replace(scp10_oa_status, " repository$", ""))
```

```{r}
write_csv(scp_sc_prep, here("scp_sc_prep.csv"))
```

```{r}
save(scp_sc_prep, file = here("scp_sc_prep.RData"))
```

```{r}
load(file=here("scp_sc_prep.RData"))
```

## Transform shared corpus table into wide format dataframe

- Replace space with semicolon
- Rename oa tags
- Split oa columns at semicolon

```{r}
scp_sc_wide <- scp_sc_prep %>%
  mutate(scp04_oa_status = str_replace_all(scp04_oa_status, " ", ";")) %>%
  mutate(scp07_oa_status = str_replace_all(scp07_oa_status, " ", ";")) %>%
  mutate(scp10_oa_status = str_replace_all(scp10_oa_status, " ", ";")) %>%
  mutate(scp04_oa_status = str_replace_all(scp04_oa_status, c("publisherfullgold" = "gold", "publisherhybridgold" = "hybrid", "publisherfree2read" = "bronze", "repositoryvor" = "green (vor)", "repositoryam" = "green (am)"))) %>%
  mutate(scp07_oa_status = str_replace_all(scp07_oa_status, c("publisherfullgold" = "gold", "publisherhybridgold" = "hybrid", "publisherfree2read" = "bronze", "repositoryvor" = "green (vor)", "repositoryam" = "green (am)"))) %>%
  mutate(scp10_oa_status = str_replace_all(scp10_oa_status, c("publisherfullgold" = "gold", "publisherhybridgold" = "hybrid", "publisherfree2read" = "bronze", "repositoryvor" = "green (vor)", "repositoryam" = "green (am)"))) %>%
  separate_wider_delim(scp04_oa_status, delim = ";", names_sep = "_", too_few = "align_start") %>%
  separate_wider_delim(scp07_oa_status, delim = ";", names_sep = "_", too_few = "align_start") %>%
  separate_wider_delim(scp10_oa_status, delim = ";", names_sep = "_", too_few = "align_start")
```

```{r}
write_csv(scp_sc_wide, here("scp_sc_wide.csv"))
```


## Transform shared corpus wide dataframe into long format

```{r}
scp_sc_long <- scp_sc_wide %>%
  pivot_longer(
    cols = contains("oa_status"),
    names_to = "snapshot",
    values_to = "oa_status",
    values_drop_na = TRUE     
  ) %>%
  mutate(snapshot = str_replace(snapshot, "_\\d$",""))
```

```{r}
write_csv(scp_sc_long, here("scp_sc_long.csv"))
```


# Prepare data for OA share calculation

```{r}
scp_oa_share <- scp_sc_prep %>%
  mutate(scp04_is_oa = ifelse(scp04_oa_status == "none", "no","yes")) %>%
  mutate(scp07_is_oa = ifelse(scp07_oa_status == "none", "no","yes")) %>%
  mutate(scp10_is_oa = ifelse(scp10_oa_status == "none", "no","yes")) %>%
  select(scp04_item_id, scp04_is_oa,scp07_is_oa,scp10_is_oa) %>%
  pivot_longer(
    cols = contains("is_oa"),
    names_to = "snapshot",
    values_to = "is_oa",
    values_drop_na = TRUE     
  ) %>%
  group_by(snapshot, is_oa) %>%
 summarise(
 n = n_distinct(scp04_item_id)
 ) %>%
 group_by(snapshot) %>%
 mutate(
 total = sum(n),
 share = n / total * 100
 ) %>%
 select(snapshot, is_oa, n, share) %>%
 mutate(snapshot = ifelse(snapshot =="scp04_is_oa", "April 24",
                          ifelse(snapshot == "scp07_is_oa", "July 24",
                                 ifelse(snapshot == "scp10_is_oa","October 24",snapshot))),
        n = formatC(n, format = "d", big.mark = ",", decimal.mark = "."),
        share = formatC(share, format = "f", digits = 2, decimal.mark = "."))
scp_oa_share
```

```{r}
write_csv(scp_oa_share, here("scp_oa_share.csv"))
```

```{r}
save(scp_oa_share, file = here("scp_oa_share.RData"))
```


# Prepare data for sankey visualisation

```{r}
scp_sn_long <- scp_sc_wide %>%
   mutate(scp04_oa_status_2 = ifelse(is.na(scp04_oa_status_2) & !is.na(scp07_oa_status_2), "missing", scp04_oa_status_2)) %>%
   mutate(scp04_oa_status_3 = ifelse(is.na(scp04_oa_status_3) & !is.na(scp07_oa_status_3), "missing", scp04_oa_status_3)) %>%
   mutate(scp07_oa_status_2 = ifelse(is.na(scp07_oa_status_2) & !is.na(scp04_oa_status_2), "missing", scp07_oa_status_2)) %>%
   mutate(scp07_oa_status_3 = ifelse(is.na(scp07_oa_status_3) & !is.na(scp04_oa_status_3), "missing", scp07_oa_status_3)) %>%
   mutate(scp07_oa_status_2 = ifelse(is.na(scp07_oa_status_2) & !is.na(scp10_oa_status_2), "missing", scp07_oa_status_2)) %>%
   mutate(scp07_oa_status_3 = ifelse(is.na(scp07_oa_status_3) & !is.na(scp10_oa_status_3), "missing", scp07_oa_status_3)) %>%
  mutate(scp10_oa_status_2 = ifelse(is.na(scp10_oa_status_2) & !is.na(scp07_oa_status_2), "missing", scp10_oa_status_2)) %>%
  mutate(scp10_oa_status_3 = ifelse(is.na(scp10_oa_status_3) & !is.na(scp07_oa_status_3), "missing", scp10_oa_status_3)) %>%
  pivot_longer(
    cols = contains("oa_status"),
    names_to = "snapshot",
    values_to = "oa_status",
    values_drop_na = TRUE     
  ) %>%
  mutate(snapshot = str_replace(snapshot, "_\\d$",""))
```

```{r}
write_csv(scp_sn_long, here("scp_sn_long.csv"))
```


# Datamanipulation for sankey visualisation

## Nodes

```{r}
scp_0407_sn <- scp_sn_long %>%
  select(scp04_item_id,snapshot, oa_status) %>%
  filter(snapshot != "scp10_oa_status") %>%
  group_by(snapshot, oa_status) %>%
  mutate(n0407 = n_distinct(scp04_item_id))

scp_0710_sn <- scp_sn_long %>%
  select(scp04_item_id,snapshot, oa_status) %>%
  filter(snapshot != "scp04_oa_status") %>%
  group_by(snapshot, oa_status) %>%
  mutate(n0710 = n_distinct(scp04_item_id))

scp_0710_sn

scp_0407_nodes_prep <- scp_0407_sn %>%
  select(-scp04_item_id) %>%
  distinct() %>%
  mutate(oa_status = ifelse(snapshot == "scp04_oa_status", toupper(oa_status), 
          ifelse(snapshot == "scp07_oa_status", str_to_title(oa_status), oa_status)))

scp_0710_nodes_prep <- scp_0710_sn %>%
  select(-scp04_item_id) %>%
  distinct() %>%
  mutate(oa_status = ifelse(snapshot == "scp07_oa_status", str_to_title(oa_status), oa_status))

scp_4710 <- bind_rows(scp_0407_nodes_prep,scp_0710_nodes_prep)
scp_4710 <- scp_4710 %>%
  select(snapshot, oa_status) %>%
  arrange(snapshot) %>%
  distinct() %>%
  ungroup()

nodes <- scp_4710 %>%
 mutate(snapshot = row_number()) %>%
 rename(name = snapshot) %>%
 mutate(name = as.character(name)) %>%
 rename(label = oa_status)
```

```{r}
write_csv(nodes, here("nodes.csv"))
```


## Links

```{r}
scp_sn_links_prep <- scp_sc_wide %>%
   mutate(scp04_oa_status_2 = ifelse(is.na(scp04_oa_status_2) & !is.na(scp07_oa_status_2), "missing", scp04_oa_status_2)) %>%
   mutate(scp04_oa_status_3 = ifelse(is.na(scp04_oa_status_3) & !is.na(scp07_oa_status_3), "missing", scp04_oa_status_3)) %>%
   mutate(scp07_oa_status_2 = ifelse(is.na(scp07_oa_status_2) & !is.na(scp04_oa_status_2), "missing", scp07_oa_status_2)) %>%
   mutate(scp07_oa_status_3 = ifelse(is.na(scp07_oa_status_3) & !is.na(scp04_oa_status_3), "missing", scp07_oa_status_3)) %>%
   mutate(scp07_oa_status_2 = ifelse(is.na(scp07_oa_status_2) & !is.na(scp10_oa_status_2), "missing", scp07_oa_status_2)) %>%
   mutate(scp07_oa_status_3 = ifelse(is.na(scp07_oa_status_3) & !is.na(scp10_oa_status_3), "missing", scp07_oa_status_3)) %>%
  mutate(scp10_oa_status_2 = ifelse(is.na(scp10_oa_status_2) & !is.na(scp07_oa_status_2), "missing", scp10_oa_status_2)) %>%
  mutate(scp10_oa_status_3 = ifelse(is.na(scp10_oa_status_3) & !is.na(scp07_oa_status_3), "missing", scp10_oa_status_3))
```


```{r}
st1 <- scp_sn_links_prep %>%
  select(scp04_item_id, scp04_oa_status_1, scp07_oa_status_1) %>%
  mutate(source = toupper(scp04_oa_status_1)) %>%
  mutate(target = str_to_title(scp07_oa_status_1)) %>%
  select(scp04_item_id, source, target)

st2 <- scp_sn_links_prep %>%
  select(scp04_item_id, scp04_oa_status_2, scp07_oa_status_2) %>%
  filter(!(is.na(scp04_oa_status_2) & is.na(scp07_oa_status_2))) %>%
  mutate(scp04_oa_status_2 = ifelse(scp07_oa_status_2 == "missing" & is.na(scp04_oa_status_2),"missing",scp04_oa_status_2)) %>%
  mutate(source = toupper(scp04_oa_status_2)) %>%
  mutate(target = str_to_title(scp07_oa_status_2)) %>%
  select(scp04_item_id, source, target)

st3 <- scp_sn_links_prep %>%
  select(scp04_item_id, scp04_oa_status_3, scp07_oa_status_3) %>%
  filter(!(is.na(scp04_oa_status_3) & is.na(scp07_oa_status_3))) %>%
  mutate(scp04_oa_status_3 = ifelse(scp07_oa_status_3 == "missing" & is.na(scp04_oa_status_3),"missing",scp04_oa_status_3)) %>%
  mutate(source = toupper(scp04_oa_status_3)) %>%
  mutate(target = str_to_title(scp07_oa_status_3)) %>%
  select(scp04_item_id, source, target)

scp0407_links_prep <- bind_rows(st1,st2,st3)
stn <- scp0407_links_prep %>%
  group_by(source, target) %>%
  summarise(n = n_distinct(scp04_item_id))


st4 <- scp_sn_links_prep %>%
  select(scp04_item_id, scp07_oa_status_1, scp10_oa_status_1) %>%
  mutate(source = str_to_title(scp07_oa_status_1)) %>%
  mutate(target = scp10_oa_status_1) %>%
  select(scp04_item_id, source, target)

st5 <- scp_sn_links_prep %>%
  select(scp04_item_id, scp07_oa_status_2, scp10_oa_status_2) %>%
  filter(!(is.na(scp07_oa_status_2) & is.na(scp10_oa_status_2))) %>%
  mutate(source = str_to_title(scp07_oa_status_2)) %>%
  mutate(target = scp10_oa_status_2) %>%
  select(scp04_item_id, source, target)

st6 <- scp_sn_links_prep %>%
  select(scp04_item_id, scp07_oa_status_3, scp10_oa_status_3) %>%
  filter(!(is.na(scp07_oa_status_3) & is.na(scp10_oa_status_3))) %>%
  mutate(source = str_to_title(scp07_oa_status_3)) %>%
  mutate(target = scp10_oa_status_3) %>%
  select(scp04_item_id, source, target)

scp0710_links_prep <- bind_rows(st4,st5,st6)
stn2 <- scp0710_links_prep %>%
  group_by(source, target) %>%
  summarise(n = n_distinct(scp04_item_id))

links <- bind_rows(stn,stn2) %>%
         mutate(source = match(source, nodes$label) -1,
                target = match(target, nodes$label) -1)
```

```{r}
write_csv(links, here("links.csv"))
```


# Sample DOIs for Unpaywall check

```{r}
doi_sample_0407 <- scp_sc_prep %>%
  select(scp04_doi, scp04_oa_status,scp07_oa_status) %>%
  filter(grepl("repositoryvor|repositoryam", scp04_oa_status) & !grepl("repositoryvor|repositoryam", scp07_oa_status)) %>%
  filter(!is.na(scp04_doi)) %>%
  filter(!duplicated(scp04_doi)) %>%
  sample_n(10000) %>%
  pull(scp04_doi)
```

```{r}
save(doi_sample_0407, file = "doi_sample_0407.RData")
```

```{r}
doi_sample_0710 <- scp_sc_prep %>%
  select(scp04_doi, scp07_oa_status,scp10_oa_status) %>%
  filter(grepl("repositoryvor|repositoryam", scp07_oa_status) & !grepl("repositoryvor|repositoryam", scp10_oa_status)) %>%
  filter(!is.na(scp04_doi)) %>%
  filter(!duplicated(scp04_doi)) %>%
  sample_n(10000) %>%
  pull(scp04_doi)
```

```{r}
save(doi_sample_0710, file = "doi_sample_0710.RData")
```
